Benefits and risks associated with children’s and adolescents’ interactions with electronic screens: An umbrella review

### 2021-09-26 reports/manuscript.Rmd
Transfer commit

---
title: "Benefits and risks associated with children’s and adolescents’ interactions with electronic screens: An umbrella review"
author:
  - name: Taren Sanders
    url: https://orcid.org/0000-0002-4504-6008
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 0000-0002-4504-6008
  - name: Philip Parker
    url: https://orcid.org/0000-0002-4604-8566 
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 0000-0002-4604-8566
  - name: Michael Noetel
    url: https://orcid.org/0000-0002-6563-8203 
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 0000-0002-6563-8203
  - name: Chris Lonsdale
    url: https://orcid.org/000-0002-2523-5565  
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 000-0002-2523-5565 
date: "`r Sys.Date()`"
repository_url: https://github.com/Motivation-and-Behaviour/screen_umbrella
bibliography: combined.bib
csl: nature.csl
always_allow_html: true
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

@@

@@

@@

```{r prisma, fig.cap="PRISMA Diagram"}
prisma$diag
```

@@

```{r}
qual_data <- reviews_table$data %>%
  filter(`Outcomes Assessed` !="<ul><li></li></ul>")

@@

```{r}
edu_effect <-   
  combined_effects %>%
  filter(outcome_category=="education" & use_effect)

@@

```{r eduplot, fig.width=plots[[1]]$dims[[1]], fig.height=plots[[1]]$dims[[2]], fig.cap="Education outcomes"}
plots[[1]]$plot
```
Among the statistically credible effects, general screen use, TV viewing, and video games were all negatively associated with academic performance or general learning. E-books that included narration, as well as touch screen education interventions, and augmented reality interventions were positively associated with learning. General screen use was also negatively associated with literacy outcomes. However, if the screen involved co-viewing (e.g., watching with a parent) or the content was educational the association with literacy was positive. Numeracy outcomes were positively associated with screen-based mathematics interventions and video games with numeracy content.

```{r}
n_edu_p999 <- edu_certain %>% filter(reanalysis_p.value<0.001) %>% nrow()
n_edu_p95 <- edu_certain %>% filter(reanalysis_p.value<0.05) %>% nrow() - n_edu_p999

@@

```{r}
health_effect <-   
  combined_effects %>%
  filter(outcome_category != "education" & use_effect)

@@

We identified `r n_health_effect` unique outcome-exposure combinations associated with health or health-related behaviour outcomes. `r `xfun::numbers_to_words(n_health_certain, cap=TRUE)`of these meta-analytic associations met our criteria for credible evidence (see Figure \@ref(fig:healthplot)). We removed `r n_health_noindiv` for not providing individual study-level data, `r n_health_samplesize` for having meta-analytic effects with samples < 1,000, and `r n_health_faileggers` had a significant Egger's test or insufficient studies to conduct the test. No remaining studies showed evidence of excessive significance. Effects not meeting one or more of these standards are presented in the supplementary material.

```{r healthplot, fig.width=plots[[2]]$dims[[1]], fig.height=plots[[2]]$dims[[2]], fig.cap="Health and health-related behaviour outcomes"}
plots[[2]]$plot
```

@@

```{r}
n_health_p999 <- health_certain %>% filter(reanalysis_p.value<0.001) %>% nrow()
n_health_p95 <- health_certain %>% filter(reanalysis_p.value<0.05) %>% nrow() - n_edu_p999

@@

# Supplementary Material

#### Full Size Forest Plots

* [Education Outcomes](https://drive.google.com/file/d/1ODsKf3jjEGNCpUTz5tju8AHRpk_xIz-e/view?usp=sharing)
* [Health-Related Outcomes](https://drive.google.com/file/d/1wdKzliWLEFuVveAj6-MzOAsoqBhnkRJW/view?usp=sharing)
* [Supplemental Education Outcomes](https://drive.google.com/file/d/1wUotZRh-jvdBJh5dkstkMtvcz7DPn-vS/view?usp=sharing)
* [Supplemental Health-Related Outcomes](https://drive.google.com/file/d/1-3frrZ3woWjPqkO84Dn3_1CCaE8thIZn/view?usp=sharing)

```{r, results='asis'}
if (knitr::pandoc_to()!="html5") "# References"
```

>>>>

---
title: "Benefits and risks associated with children’s and adolescents’ interactions with electronic screens: An umbrella review"
author:
  - name: Taren Sanders
    url: https://orcid.org/0000-0002-4504-6008
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 0000-0002-4504-6008
  - name: Philip Parker
    url: https://orcid.org/0000-0002-4604-8566 
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 0000-0002-4604-8566
  - name: Michael Noetel
    url: https://orcid.org/0000-0002-6563-8203 
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 0000-0002-6563-8203
  - name: Chris Lonsdale
    url: https://orcid.org/000-0002-2523-5565  
    affiliation: Australian Catholic University
    affiliation_url: https://www.acu.edu.au/research/our-research-institutes/institute-for-positive-psychology-and-education
    orcid_id: 000-0002-2523-5565 
date: "`r Sys.Date()`"
repository_url: https://github.com/Motivation-and-Behaviour/screen_umbrella
bibliography: combined.bib
csl: nature.csl
always_allow_html: true
output:
  github_document:
    html_preview: false
    pandoc_args: "--wrap=none"
---

@@

render_img <- knitr::pandoc_to()=="html5"
inc_ref <- knitr::pandoc_to()!="html5"

@@

Philip.Parker@acu.edu.au will insert an explanation about why he forced us to move away from our original [evidence assessment plan](https://www.crd.york.ac.uk/prospero/display_record.php?RecordID=76051).

@@

```{r prisma, eval=render_img, fig.cap="PRISMA Diagram"}
prisma$diag
```

@@

```{r quality}
qual_data <- reviews_table$data %>%
  filter(`Outcomes Assessed` !="<ul><li></li></ul>")

@@

```{r eduoutcomes}
edu_effect <-   
  combined_effects %>%
  filter(outcome_category=="education" & use_effect)

@@

```{r eduplot, eval=render_img, fig.width=plots[[1]]$dims[[1]], fig.height=plots[[1]]$dims[[2]], fig.cap="Education outcomes"}
plots[[1]]$plot
```

Among the statistically credible effects, general screen use, TV viewing, and video games were all negatively associated with academic performance or general learning. E-books that included narration, as well as touch screen education interventions, and augmented reality interventions were positively associated with learning. General screen use was also negatively associated with literacy outcomes. However, if the screen involved co-viewing (e.g., watching with a parent) or the content was educational the association with literacy was positive. Numeracy outcomes were positively associated with screen-based mathematics interventions and video games with numeracy content.

```{r educertain}
n_edu_p999 <- edu_certain %>% filter(reanalysis_p.value<0.001) %>% nrow()
n_edu_p95 <- edu_certain %>% filter(reanalysis_p.value<0.05) %>% nrow() - n_edu_p999

@@

```{r healthoutcome}
health_effect <-   
  combined_effects %>%
  filter(outcome_category != "education" & use_effect)

@@

We identified `r n_health_effect` unique outcome-exposure combinations associated with health or health-related behaviour outcomes. `r xfun::numbers_to_words(n_health_certain, cap=TRUE)` of these meta-analytic associations met our criteria for credible evidence (see Figure \@ref(fig:healthplot)). We removed `r n_health_noindiv` for not providing individual study-level data, `r n_health_samplesize` for having meta-analytic effects with samples < 1,000, and `r n_health_faileggers` had a significant Egger's test or insufficient studies to conduct the test. No remaining studies showed evidence of excessive significance. Effects not meeting one or more of these standards are presented in the supplementary material.

```{r healthplot, eval=render_img, fig.width=plots[[2]]$dims[[1]], fig.height=plots[[2]]$dims[[2]], fig.cap="Health and health-related behaviour outcomes"}
plots[[2]]$plot
```

@@

```{r healthcertain}
n_health_p999 <- health_certain %>% filter(reanalysis_p.value<0.001) %>% nrow()
n_health_p95 <- health_certain %>% filter(reanalysis_p.value<0.05) %>% nrow() - n_edu_p999

@@

**Note:** *Below are rough notes on ideas for the discussion. Please feel free to add ideas. We will revise once we agree on the rest of the paper.*

#### Broad summary of the findings relating back to the purpose
When meta-analysis examined general screen use, and did not specify the content, context or device, there was strong evidence showing potentially harmful associations with academic performance, literacy, body composition, and depression. However, when meta-analyses included a more nuanced examination of exposures a more complex picture appeared. Many exposures were likely neither universally harmful nor consistently beneficial.

#### TV
Strong evidence indicated that watching television programs was associated with poorer academic performance and learning. However, there was also weak evidence that if the child watched these programs with a parent (i.e., coviewing) or the content of the program was educational then television viewing was associated with better literacy.

Outside of educational outcomes, there was strong evidence of a harmful association with body composition and marginal associations with sleep duration. Overall, there was credible evidence indicating that TV watching was likely to confer benefits for children, but it may not be harmful as long as children watch with their parents or the content is educational.

#### Video games 
* Weak evidence of harm for body comp.
* Strong evidence of harm for academic performance.
* Strong evidence of benefit for numeracy games on numeracy.

#### Screen-based Interventions
One category of exposure appeared to consistently confer benefits: Interventions delivered via screens. This finding indicates that interventions can be effectively delivered using electronic media platforms. It does not necessarily indicate that screens are more effective than other methods (e.g., face-to-face, printed material).

* Touch screen = strong evidence of benefit for learning
* Math interventions = strong evidence on numeracy.
* Augmented reality on learning, strong evidence.
* Screen-based interventions. Strong effect on health behaviours.

#### Consistent Harms
The exposures that were consistently associated with harms: social media and food advertising via screen media.  
Social media showed strong evidence of harmful associations with risk taking in general, as well as unsafe sex and substance abuse.  
Strong evidence also indicated screen-based food advertising (in general and in advergames), was risky as it was positively associated with food intake.  
Finally, there was strong evidence that sexy media was associated with sexual activity.

#### Implications of findings for targeted policy and practice
Screen time is complex. We need to stop thinking about it in simplistic terms as “bad” and something that should be minimised in all instances. Need to think about how it can best be used. For example, recreational television and video games appear to confer few, if any benefits, and appear to be associated with poor body composition and academic performance. However, tv programs and video games designed to promote learning may be effective. The current literature indicates that social media is likely harmful, but if social media platforms specifically designed to promote learning can be designed and popularised, perhaps they could have benefits.

#### Survey of the field and implications for future research

Screen time research is big and varied and rapidly growing. Reviews tended to be quite general (e.g., all screentime) and even when more targeted (e.g., social media) nuances related to specific content (e.g., Instagram vs Facebook) have not been meta-analysed.

There is credible evidence for X, we need more evidence for Y.  
All observational research with self-report evidence.

#### Limitations of the review

Can’t look at fine-grained detail - eg moderators, but we did look at different exposures and outcomes (quasi moderators).  
GRADE - we did not rate the risk of bias of individual studies in the meta-analyses. Total k  >3,000 (intractable task)  
We made decisions to present effects that reach certain thresholds (e.g., N>1,000, non-significant Egger’s test) but also provide more complete results in the supplementary files for those who wish to investigate.

#### Conclusions

# Appendix

#### Figures

* [Prisma Diagram](https://drive.google.com/file/d/1ezgeAr2WPhy6MNSESS8bhg3VMOrLAUC7/view?usp=sharing)
* [Education Outcomes](https://drive.google.com/file/d/1ODsKf3jjEGNCpUTz5tju8AHRpk_xIz-e/view?usp=sharing)
* [Health-Related Outcomes](https://drive.google.com/file/d/1wdKzliWLEFuVveAj6-MzOAsoqBhnkRJW/view?usp=sharing)
* [Supplemental Education Outcomes](https://drive.google.com/file/d/1wUotZRh-jvdBJh5dkstkMtvcz7DPn-vS/view?usp=sharing)
* [Supplemental Health-Related Outcomes](https://drive.google.com/file/d/1-3frrZ3woWjPqkO84Dn3_1CCaE8thIZn/view?usp=sharing)

#### Tables

#### Links

* [Manuscript Google Doc](https://docs.google.com/document/d/1NN4yJWQVepNjbX0mND4aidWWXcCDkg-eMepRnWdtQ9o/edit#)
* [Online Report](https://motivation-and-behaviour.github.io/screen_umbrella/)

```{r refs, eval=inc_ref, results='asis'}
"# References"
```
